Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-02 ind_files/04_unification_discussion.tex
Add discussion material

\epigraph{\justifying The grand aim of all science [is] to cover the greatest number of empirical facts by logical deduction from the smallest number of hypotheses or axioms}{Albert Einstein}
%\epigraph{\justifying Not only is algebraic reasoning exact; it imposes an exactness on the verbal postulates made before algebra can start which is usually lacking in the first verbal formulations of scientific principles.}{J.B.S. Haldane}
\justifying
In this thesis, we have seen how stochastic birth-death processes can be used to construct and analyze mechanistic individual-based models for the dynamics of finite populations. In doing so, we have also seen that various well-known equations of evolutionary dynamics can be recovered in the infinite population size limit. In the finite-dimensional case (corresponding to discrete trait variants), the infinite population limit corresponds to the equations of population genetics and evolutionary game theory. In the infinite-dimensional case (corresponding to quantitative traits), we instead obtain the equations of quantitative genetics, and, in some further limits, adaptive dynamics. In both cases, the mean value of the trait in the population changes according to an equation resembling the Price equation. My derivatioon highlights the natural connections between the various equations of population dynamics - For example, the same procedures that lead to the replicator-mutator equation in the case of discretely varying traits yield Kimura's model in the quantitative case, underscoring the broad similarities between evolutionary game theory and quantitative genetics. The major formulations are summarized in Table \ref{table_summary}.

@@

\section{The fundamental theorem for changes in type frequencies in the population}\label{sec_fun_theorem_freq}

@@

\section{The fundamental theorem for the mean value of a type-level quantity in the population}\label{sec_fun_theorems_mean}

@@

\section{The fundamental theorem for the variance of a type-level quantity in the population}\label{sec_fun_theorems_var}

@@

\section{Weak-Noise approximation}
If the population is large (but finite) and the stochasticity is sufficiently weak, stochastic dynamics can be studied analytically using the weak noise approximation. Usually, this approximation is valid if we are studying a stochastic trajectory that is fluctuating about a point that is stable in the deterministic limit \citep{van_kampen_stochastic_1981}. Such situations occur often, since many systems of the forms studied here quickly relax to stable equilibria.

\section{Towards a stochastic evolutionary theory}

One striking feature that repeatedly shows up in our derivations is that finite populations exhibit phenomena that are not visible in infinite population models. For example, in both the stochastic logistic equation \ref{ex_1D_stoch_logistic_BD_eqns} and in two-strategy games with finite population sizes \citep{tao_stochastic_2007}, demographic noise ensures that all populations are guaranteed to go extinct given enough time, even if the deterministic limit predicts a stable state far from extinction. In the case of quantitative traits, demographic noise can hinder adaptive diversification by increasing the time before evolutionary branching occurs \citep{claessen_delayed_2007, wakano_evolutionary_2013, debarre_evolutionary_2016}, causing stochastic extinction of existing evolutionary branches \citep{rogers_demographic_2012, johansson_will_2006}, or preventing branching altogether if the population is too small \citep{rogers_modes_2015, johnson_two-dimensional_2021}. Stochastic systems also routinely exhibit evolution towards attractors that cannot be attained in the deterministic limit \citep{delong_stochasticity_2023}, sometimes even completely reversing the direction of evolution predicted by deterministic dynamics \citep{constable_demographic_2016,mcleod_social_2019}. Since real-life populations are stochastic and finite, it is thus imperative that modellers work with stochastic first-principles models instead of their deterministic limits, lest they risk missing important phenomena that are unique to stochastic systems \citep{black_stochastic_2012,schreiber_does_2022,hastings_transients_2004,shoemaker_integrating_2020}. In the context of our models, we have seen that if we observe the change in trait frequencies instead of the change in densities, finite populations are subject to an additional evolutionary force that vanishes in infinite population models.

>>>>

\epigraph{\justifying The grand aim of all science [is] to cover the greatest number of empirical facts by logical deduction from the smallest number of hypotheses or axioms}{Albert Einstein}
%\epigraph{\justifying Not only is algebraic reasoning exact; it imposes an exactness on the verbal postulates made before algebra can start which is usually lacking in the first verbal formulations of scientific principles.}{J.B.S. Haldane}
\justifying
In this thesis, we have seen how stochastic birth-death processes can be used to construct and analyze mechanistic individual-based models for the dynamics of finite populations. In doing so, we have also seen that various well-known equations of evolutionary dynamics can be recovered in the infinite population size limit. In the finite-dimensional case (corresponding to discrete trait variants), the infinite population limit corresponds to the equations of population genetics and evolutionary game theory. In the infinite-dimensional case (corresponding to quantitative traits), we instead obtain the equations of quantitative genetics, and, in some further limits, gradient dynamics. In both cases, the mean value of the trait in the population changes according to an equation resembling the Price equation. My derivation highlights the natural connections between the various equations of population dynamics - For example, the same procedures that lead to the replicator-mutator equation in the case of discretely varying traits yield Kimura's model in the quantitative case, underscoring the broad similarities between evolutionary game theory and quantitative genetics and extending known similarities to the finite population case. The major formulations are summarized in Table \ref{table_summary}.

@@

\section{Fundamental theorems of evolution in finite population}
\subsection{The fundamental theorem for changes in type frequencies in the population}\label{sec_fun_theorem_freq}

@@

\subsection{The fundamental theorem for the mean value of a type-level quantity in the population}\label{sec_fun_theorems_mean}

@@

\subsection{The fundamental theorem for the variance of a type-level quantity in the population}\label{sec_fun_theorems_var}

@@

\section{Discussion \& Outlook}\label{sec_disc}

Stochastic finite population models often exhibit behaviors that are markedly different from their deterministic limits. Since real-life populations are stochastic and finite, it is thus imperative that modellers work with stochastic first-principles models instead of their deterministic limits, lest they risk missing important phenomena that are unique to stochastic systems \citep{black_stochastic_2012,schreiber_does_2022,hastings_transients_2004,shoemaker_integrating_2020}. In the context of eco-evolutionary theory, \cite{doebeli_towards_2017} have called for a reformulation of evolutionary dynamics starting from stochastic birth-death processes on the grounds that such a formulation is more fundamental and mechanistic, a view that has also been echoed by theorists before \citep{metcalf_why_2007,geritz_mathematical_2012}. In this thesis, I present and derive from first principles some fundamental relations (equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance}) that any such reformulation must satisfy. Further, these relations deal with biologically important quantities, lend themselves to simple biological interpretation, and are very general. They thus fulfill the criteria to be called `fundamental theorems' (in the sense of \cite{queller_fundamental_2017}), or `unifying principles' (in the sense of \cite{lion_theoretical_2018}) for the dynamics of finite populations. \cite{lion_theoretical_2018} has pointed out that in the dynamic setting (for infinite populations), the replicator-mutator equation \eqref{nD_replicator_mutator} is in some sense the `most fundamental' of the lot, and equations like the Price equation are best viewed as a hierarchy of moment equations for the population mean, population variance, etc. of a type-level quantity. This is also true in our framework - Equation \eqref{nD_stochastic_RM} is the most fundamental equation for population dynamics, and equations like \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} can then be derived from \eqref{nD_stochastic_RM} through repeated application of It\^o's formula (in principle for any moment of the distribution of $f$ in the population, though this quickly becomes too tedious to actually carry out in practice). Just like in infinite population models, the various terms of equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance} lend themselves to simple biological interpretation. Note that actually solving these equations analytically for equilibrium/stationary state distributions of $\mathbf{p}$, $\overline{f}$, and $\sigma^2_f$ will quickly become impossibly difficult if the birth and death rate functions are complicated. Indeed, previous studies indicate that for $m > 2$, birth-death models can exhibit a dizzying array of complicated phenomena, including limit cycles and evolutionary chaos \citep{doebeli_diversity_2017}. Nevertheless, our equations will still hold for arbitrarily complicated birth and death rates. Like the Price equation, the utility of these equations lies not in their solutions, but instead in their generality and the fact that their terms help us clearly understand the various forces operating in biological populations \citep{frank_natural_2012, luque_mirror_2021}. The general approach of working with `high-level' processes without specifying system-specific details is part of a broader pursuit of `model-independent' eco-evolutionary theory that has recently been gaining popularity in the literature  \citep{queller_fundamental_2017, doebeli_towards_2017, lion_theoretical_2018, allen_mathematical_2019, rice_universal_2020, week_white_2021, wickman_theoretical_2022, kuosmanen_turnover_2022, mazzolini_universality_2022}. Such model-independent descriptions that `abstract away' system-specific details almost inevitably come at the cost of precision \citep{levins_strategy_1966, potochnik_idealization_2018}, and are thus intended to complement rather than supersede specific models of specific systems. In our case, our framework complements the various specific studies of social evolution in specific models \citep{constable_demographic_2016, houchmandzadeh_selection_2012, chotibut_evolutionary_2015, mcleod_social_2019} that pointed to noise-induced selection reversing the direction of evolution predicted by deterministic models.\\
Noise-induced selection due to variation in per-capita turnover rate $\tau$ is an evolutionary force only seen in finite populations. Mathematically, this term arises essentially due to It\^o's formula adding an extra term when conducting a change of variables in going from type densities to type frequencies. Biologically, it arises due to different types experiencing a different number of stochastic events (birth and death) in a given time interval. The effects of noise-induced selection are increasingly being recognized in various specific models \citep{houchmandzadeh_selection_2012, houchmandzadeh_fluctuation_2015, constable_demographic_2016, parsons_pathogen_2018}, and, recently, have also been shown in some more general settings \citep{week_white_2021,mazzolini_universality_2022,kuosmanen_turnover_2022}. This force has some important practical implications. For one, simulation studies that work with evolutionary individual-based or agent-based models should be careful about whether interactions effects are incorporated into birth rates or death rates, since this seemingly arbitrary choice can have unintended consequences due to noise-induced selection \citep{mcleod_social_2019,kuosmanen_turnover_2022}. Similarly, being mindful of noise-induced selection is important for conservation and population management, two applied fields that regularly deal with small populations (\emph{i.e.} populations which likely experience stronger noise-induced selection). For example, when trying to increase the population of a hypothetical desired species in a multispecies community, increasing the birth rate is \emph{not} equivalent to reducing the death rate even though both result in an increase in the Malthusian fitness (growth rate) $w_i$ - Decreasing the death rate leads to a decrease in $\tau_i$, which leads to positive noise-induced selection (from \eqref{nD_stochastic_RM}), whereas increasing the birth rate leads to an increase in $\tau_i$, which leads to noise-induced selection acting to reduce the abundance of the focal species from the community. If the total community size is small, increasing the birth rate of a species can thus lead to noise-induced selection completely eliminating the focal species from the community despite the fact that our actions \emph{increased} the growth rate of this species. Lastly, equations \eqref{nD_stochastic_RM} and \eqref{nD_stochastic_Price} imply that for evolution to be \emph{truly} neutral in finite populations, it is not sufficient for the trait in question to be neutral with respect to fitness $w$. Instead, we also require the trait to be neutral with respect to turnover rate $\tau$. A similar phenomenon has also been recognized in some finite population ecological models, where seemingly neutral models (equal growth rate) nevertheless select for individuals with greater longevity \citep{lin_features_2012, oliveira_advantage_2017} \\
At first glance, the idea that individuals with lower birth and death rates are preferred over individuals with higher birth and death rates may be reminiscent of ideas like $r$ vs $K$ selection in life-history evolution \citep{stearns_evolution_1977}. However, it is unclear whether this similarity reflects some deep principle or whether it is just superficial. Models in life-history theory are often primarily concerned with temporally fluctuating (external) environments, a factor that I have neglected entirely in my formalism. In principle, it is possible to make the birth and death rates \ref{nD_functional_forms_for_replicator} in our model also depend on a temporally varying external environment $E(t)$ (whose rules of variation may possibly depend on the population $\mathbf{v}$). Incorporating such a term would ensure that the `ecological feedback' terms in equations \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} are non-zero, but may also lead to much more complex dynamics. Furthermore, if the environment $E(t)$ has stochasticity, the system as a whole would have two qualitatively different forms of noise --- \emph{extrinsic} noise from the environment, and \emph{intrinsic} noise from the finiteness of the population --- interacting with each other, making the dynamics complicated and likely analytically intractable. Ecological frameworks such as modern coexistence theory, which deal with similar questions about population dynamics as our model and would benefit from a first principles birth-death approach, also generally work with fluctuating external environments \citep{chesson_multispecies_1994}. As such, while the idea of integrating our birth-death framework with ecological ideas such as the pace-of-life syndrome \citep{mathot_models_2018, wright_life-history_2019} or modern coexistence theory \citep{chesson_multispecies_1994, johnson_resolving_2022} is biologically appealing, it is far from trivial and may present a promising avenue for future studies.
